Analyse the **`r ONT` hierarchy** of Gene Ontology.

```{r}
# create sublist for over-enrichment analysis
LIST_LEVEL <- paste0(ONT, "-OE")
enrichment_list[[LIST_LEVEL]] <- list()
number_genes_enrichment[[LIST_LEVEL]] <- list()
# fill the sublist with over-enrichment results
for (i in 1:length(genes_list_arab))
{
  if (ncol(genes_list_arab[[1]])==1) {
    col <- 1
  } else {
    col <- 2
  }
  temp <- enrichGO(gene = genes_list_arab[[i]][[col]],
                   OrgDb = organism,
                   keyType = KEYTYPE,
                   ont = ONT,
                   pvalueCutoff = P_CUTOFF,
                   qvalueCutoff = Q_CUTOFF,
                   pAdjustMethod = P_ADJUST_METHOD,
                   minGSSize = MIN_GSSIZE,
                   maxGSSize = MAX_GSSIZE,
                   readable = FALSE,
                   pool = FALSE)
  
  # check if there was enrichment
  empty <- length(temp$ID) == 0

  # if it's empty we don't save it in the list
  if (empty == FALSE)
  {
    # save values in the list
    name <- names(genes_list_arab)[i]
    enrichment_list[[LIST_LEVEL]][[name]] <- temp
    # save number of genes analysed in a vector
    number_genes_enrichment[[LIST_LEVEL]][[name]] <- length(genes_list_arab[[i]][[col]])
  } else {
    message(paste0("0 enrichment terms found for ", names(genes_list_arab)[i]))
  }
}

rm(temp)
```

Remove empty list elements, if any.

```{r}
#remove empty list(s) before plot
length_loop <- length(enrichment_list)
for (i in 1:length_loop) {
  if (length(enrichment_list[[i]]) == 0) {
    enrichment_list[[i]] <- NULL
    length_loop <- length_loop - 1
    }
}
```

Assign colors to the TF for the ggplot.

```{r, eval=EVAL_TF, include=EVAL_TF}
# edit tf variable to assign colors
tf_edited <- tf$Gene_ID
tf_edited <- sort(tf_edited)
tf_edited <- unique(tf_edited)
tf_edited <- as.data.frame(tf_edited)
tf_edited$type <- "tf"
colnames(tf_edited)[1] <- "ID"
# merge tf and other_genes_arab to assign colors
if (exists("tf_edited") & exists("other_genes_arab")) {
  tf_and_other <- rbind(tf_edited, other_genes_arab)
  # get a vector of repeated IDs
  repeated <- data.frame(table(tf_and_other$ID))
  repeated <- repeated[repeated$Freq > 1, ]
  repeated <- as.vector(repeated$Var1)
  # remove them from the dataframe
  if (length(repeated) > 0) {
    tf_and_other <- tf_and_other[!tf_and_other$ID %in% repeated, ]
    # add them again
    repeated <- as.data.frame(repeated)
    repeated$type <- "both"
    colnames(repeated)[1] <- "ID"
    tf_and_other <- rbind(tf_and_other, repeated)
  }
} else tf_and_other <- tf_edited
# assign colors
ggcolors <- list()
ggcolors[[LIST_LEVEL]] <- list()
for (i in 1:length(enrichment_list[[LIST_LEVEL]])) {
  # keep result field for each element in the list
  tmp <- enrichment_list[[LIST_LEVEL]][[i]]@result
  # add new column to save TF results
  enrichment_list[[LIST_LEVEL]][[i]]@result$TF <- NA
  name <- names(enrichment_list[[LIST_LEVEL]][i])
  ggcolors[[LIST_LEVEL]][[name]] <- list()
  for (j in 1:length(rownames(tmp))) {
    # get j geneID and compare it with the TF list
    tmp2 <- tmp[j,]$geneID
    tmp2 <- unlist(strsplit(tmp2, "/"))
    # assign TFs and add them to the enrichment object
    tmp3 <- merge(x = as.data.frame(tmp2),
                  y = tf,
                  by.x = 1,
                  by.y = 2)
    tmp3 <- unique(tmp3[, -2])
    tmp3 <- tmp3$Family
    tmp3 <- tmp3[order(tmp3)]
    # add those TF to the enrichment object
    if (!is_empty(tmp3)) {
      tmp3 <- paste0(tmp3, collapse = "|") # we use | instead of / because one of the TF's name has a /
      enrichment_list[[LIST_LEVEL]][[i]]@result$TF[[j]] <- tmp3
      } # end if
    # assign colors for the ggplot
    tmp4 <- merge(x = as.data.frame(tmp2),
                  y = tf_and_other,
                  by = 1)
    tmp4 <- tmp4$type
    # assign pink if a process has both TF and other genes
    if (any(tmp4 == "both") | (any(tmp4 == "tf") && any(tmp4 == "other_genes"))) {ggcolors[[LIST_LEVEL]][[i]] <- append(ggcolors[[LIST_LEVEL]][[i]], setNames("Both categories", tmp[j,]$Description))
      } else { # assign orange if it's TF
        if (any(tmp4 == "tf")) {
          ggcolors[[LIST_LEVEL]][[i]] <- append(ggcolors[[LIST_LEVEL]][[i]], setNames("Transcription factors", tmp[j,]$Description))
          } else { # assign purple if it's other gene
            if (any(tmp4 == "other_genes")) {
              ggcolors[[LIST_LEVEL]][[i]] <- append(ggcolors[[LIST_LEVEL]][[i]], setNames("Genes of interest", tmp[j,]$Description))
            } else { # assign green if none of the above
              ggcolors[[LIST_LEVEL]][[i]] <- append(ggcolors[[LIST_LEVEL]][[i]], setNames("None", tmp[j,]$Description))
              } # end else purple
            } # end else orange
      } # end else pink
    # names(ggcolors[[LIST_LEVEL]][[i]][j]) <- tmp[j,]$Description # this doesn't work
    } # end for j
  } # end for i
```

```{r, eval=(EVAL_TF && ONT=="BP"), child='11_TF.Rmd'}
```

### Dotplot

```{r, fig.show = "hold", out.width = "50%", fig.height=0.12*ENRICH_FACTOR, fig.width=5}
for (i in 1:length(enrichment_list[[LIST_LEVEL]])) {
  tmp <- enrichment_list[[LIST_LEVEL]][[i]]
  MIN_GENES <- trunc(number_genes_enrichment[[LIST_LEVEL]][[i]]/ENRICH_FACTOR)
  if (MIN_GENES == 0) {
    MIN_GENES <- 1
  } # end if
  tmp2 <- tmp$Count >= MIN_GENES
  # if no process passes the filter, we skip the plots
  if (!any(tmp2)) {
    next
  } # end if
  tmp <- tmp[tmp2]$Description
  
  # dotplot
  set.seed(SEED)
  p_dp <- dotplot(enrichment_list[[LIST_LEVEL]][[i]],
                    showCategory = tmp, # seleccionamos cuántos procesos
                    font.size = 6,
                    label_format = 80) + # sets wrap length
      ggtitle(paste0(names(enrichment_list[[LIST_LEVEL]][i]), " enriched GOs (", ONT, ")")) +
      theme(plot.title = element_text(hjust = 0.9))
    
    print(p_dp)
}
```

### Treeplot

```{r}
warning("Be careful with the interpretation of treeplots: if there are more than 10 clusters, the processes might be wrong placed")
```

```{r, fig.show = "hold", out.width = "50%", fig.height=0.15*ENRICH_FACTOR, fig.width=18, warning=FALSE, message=FALSE, eval=LIST_LEVEL=="BP-OE", include=LIST_LEVEL=="BP-OE"}
# VOY A REPRESENTAR LOS TREEPLOT SOLO CON BP-OE PORQUE CON CC Y MF FALLA. HAY QUE ARREGLARLO
for (i in 1:length(enrichment_list[[LIST_LEVEL]])) {
  tmp <- enrichment_list[[LIST_LEVEL]][[i]]
  MIN_GENES <- trunc(number_genes_enrichment[[LIST_LEVEL]][[i]]/ENRICH_FACTOR)
  if (MIN_GENES == 0) {
    MIN_GENES <- 1
  } # end if
  tmp2 <- tmp$Count >= MIN_GENES
  # if no process passes the filter, we skip the plots
  if (!any(tmp2)) {
    next
  } # end if
  enrichment_filt <- tmp[tmp2]
  tmp <- enrichment_filt$Description
  
  # treeplot
  set.seed(SEED)
  tmp_tree <- pairwise_termsim(enrichment_list[[LIST_LEVEL]][[i]],
                               showCategory = nrow(enrichment_list[[LIST_LEVEL]][[i]]))
  # to get the number of clusters
  if (NCLUST == 0){
    mat <- tmp_tree@termsim # ONLY HAS UPPER TRIANGLE
    tmat <- t(mat)
    mat[lower.tri(mat)] <- tmat[lower.tri(tmat)] # complete lower triangle
    mat[is.na(mat)] <- 1 # set diagonal to 1
    mat_dist <- as.dist(mat)
    mat_tree <- hclust(mat_dist, method = "ward.D")
    group_dynamic <- cutreeDynamic(dendro = mat_tree,
                                   distM = as.matrix(mat_dist),
                                   minClusterSize = 20)
    names(group_dynamic) <- mat_tree$labels
    dynamic_k <- length(table(group_dynamic))
    NCLUST <- dynamic_k
  }
  # treeplot
  p_tp <- treeplot(tmp_tree,
                   showCategory = length(tmp),
                   fontsize = 3,
                   nCluster = NCLUST,
                   # label_format_tiplab = 35,
                   cluster.params = list(label_words_n = 5, label_format = 30)) +
    ggtitle(paste0(names(enrichment_list[[LIST_LEVEL]][i]), " enriched GOs (", ONT, ")")) +
    theme(plot.title = element_text(hjust = 0.9))
  print(p_tp)
}
```

### Network

```{r, fig.show = "hold", out.width = "50%", fig.height=6, fig.width=6}
network_plots <- list()
network_plots[[LIST_LEVEL]] <- list()
for (i in 1:length(enrichment_list[[LIST_LEVEL]])) {
  tmp <- enrichment_list[[LIST_LEVEL]][[i]]
  MIN_GENES <- trunc(number_genes_enrichment[[LIST_LEVEL]][[i]]/ENRICH_FACTOR)
  if (MIN_GENES == 0) {
    MIN_GENES <- 1
  } # end if
  tmp2 <- tmp$Count >= MIN_GENES
  # if no process passes the filter, we skip the plots
  if (!any(tmp2)) {
    next
  } # end if
  tmp <- tmp[tmp2]$Description
  
  # network
  set.seed(SEED)
  p_nt <- emapplot(pairwise_termsim(enrichment_list[[LIST_LEVEL]][[i]]),
                     showCategory = length(tmp), # seleccionamos cuántos procesos
                     # cex_label_category = 0.5, # tamaño etiquetas nodos
                     node_label = "none") + # quitar etiquetas 
            ggtitle(paste0(names(enrichment_list[[LIST_LEVEL]][i]), " enriched GOs (", ONT, ")"))
    
    print(p_nt)
    
    network_plots[[LIST_LEVEL]][[names(enrichment_list[[LIST_LEVEL]][i])]] <- p_nt
}
```

### Interactive network

```{r, fig.show = "hold", out.width = "50%", fig.height=6, fig.width=6}
plt <- htmltools::tagList()
plt[[LIST_LEVEL]] <- htmltools::tagList()

if (exists("ggcolors")) {
  # legend_colors <- c("Transcription factors" = "orange",
  #                    "Genes of interest" = "purple",
  #                    "Both categories" = "yellow",
  #                    "None" = "green")
for (i in 1:length(network_plots[[LIST_LEVEL]])) {
  # keep network_plots data
  tmp <- network_plots[[LIST_LEVEL]][[i]]$data
  tmp <- tmp[, 1:3]
  # keep ggcolors data
  tmp2 <- as.data.frame(unlist(ggcolors[[LIST_LEVEL]][[i]]))
  # merge both of them
  tmp3 <- merge(x = tmp,
                y = tmp2,
                by.x = 3,
                by.y = "row.names")
  colnames(tmp3)[4] <- "categories"
  set.seed(SEED)
  p_nt_int <- ggplot(tmp3,
                     aes(x = x, y = y, name = name)) +
      geom_point(aes(color = categories), size = 2) +
    scale_color_manual(name = "Processes including:",
                       breaks = c("Transcription factors", "Genes of interest", "Both categories", "None"),
                       values = c("Transcription factors" = "orange", "Genes of interest" = "purple", "Both categories" = "yellow", "None" = "green")) +
    # scale_color_manual(values = unlist(ggcolors[[LIST_LEVEL]][[i]]),
    #                    limits = force) + # define colors with ggcolors
    # theme(legend.position = "bottom") + # it doesn't work with ggplotly
      ggtitle(paste0(names(network_plots[[LIST_LEVEL]][i]), " enriched GOs (", ONT, ")"))
    plt[[LIST_LEVEL]][[i]] <- as_widget(ggplotly(p_nt_int,
                                          height = 600,
                                          width = 600))
}
} else {
  for (i in 1:length(network_plots[[LIST_LEVEL]])) {
    set.seed(SEED)
    p_nt_int <- ggplot(network_plots[[LIST_LEVEL]][[i]]$data,
                       aes(x = x, y = y, name = name)) +
      geom_point(aes(color = "red"), size = 2) +
      theme(legend.position = "none") +
      ggtitle(paste0(names(network_plots[[LIST_LEVEL]][i]), " enriched GOs (", ONT, ")"))
    plt[[LIST_LEVEL]][[i]] <- as_widget(ggplotly(p_nt_int,
                                                 height = 600,
                                                 width = 600))
  }
}
```

```{r}
message(cat("orange = transcription factor\npurple = gene of interest\nyellow = both transcription factor and gene of interest\ngreen = none of the above"))
```

```{r, echo=FALSE, fig.show = "hold", out.width = "50%", fig.height=6, fig.width=6}
plt
```

### Wordclouds by words

```{bash, engine.opts='-l', results='hide'}
# remove wordclouds directory
rm -r wordclouds
# create wordclouds directory
mkdir wordclouds
```

Get stopwords.

```{bash, engine.opts='-l', results='hide', eval=FALSE, include=FALSE}
# Obtain stopwords
curl -sL "https://raw.githubusercontent.com/stopwords-iso/stopwords-en/master/stopwords-en.txt" |
# sort stopwords and save as "stopwords"
sort | 
# save stopwords in the file stopwords-en using tee instead of ">"
tee ../stopwords-en | 
# show the first 4 words
head -n 4
```

Create the text for the wordcloud graph.

```{r}
for (i in 1:length(enrichment_list[[LIST_LEVEL]])) {
  # create data.frame with GO term frequency
  wcdf <- read.table(text = enrichment_list[[LIST_LEVEL]][[i]]$GeneRatio,
                     sep = "/")[1]
  names(wcdf)[1] <- "GeneRatio"
  # add GO term column
  wcdf$term <- enrichment_list[[LIST_LEVEL]][[i]][,2]
  # convert wcdf to text
  mytext <- c()
  mytext <- append(rep(wcdf$term, wcdf$GeneRatio), mytext)
  mytext <- paste(mytext, collapse = " ")
  mytext <- toString(mytext)
  
  # remove punctuation signs
  mytext <- gsub("[[:punct:]]", "", mytext)
  
  # save it to work with bash
  write.table(mytext,
              file = paste0("wordclouds/",
                            names(enrichment_list[LIST_LEVEL]),
                            "_",
                            names(enrichment_list[[LIST_LEVEL]][i]),
                            ".tsv"))
  }
```

EXPLICAR

```{bash, message = FALSE, results='hide'}
for file in wordclouds/*.tsv;
do
cat $file |
# converting text to lowercase
tr '[:upper:]' '[:lower:]' |
# extract words and put the on separate line
grep -oE "[a-z\']{2,}" |
# sort in alphabetical order
sort |
# select non-matching lines
grep -Fvwf ../stopwords-en-edited |
# remove all the duplicates and count frequency
uniq -c |
# sort the list by count
sort -nr |
# save the file
tee $file.wc |
# show top 4 lines
head -n 4
# remove $file
rm $file
done
```

EXPLICAR

```{r, fig.show = "hold", out.width = "50%", fig.height=6, fig.width=6}
# load all the variables for wc
filenames <- list.files(paste0(SOURCE_DIR, "Rmd/wordclouds"),
                        pattern = "*.wc",
                        full.names = TRUE)
ldf <- lapply(filenames, read.table)

filenames <- gsub(".*/", "", filenames)
filenames <- gsub(".tsv.wc", "", filenames)
names(ldf) <- filenames

for (i in 1:length(ldf)) {
  words_df <- ldf[[i]]
  words_df <- words_df[, c(2,1)]
  # MIN_FREQ <- max(words_df[[2]])/MIN_RATIO
  set.seed(SEED)
  wordcloud(words = words_df[[1]],
            freq = words_df[[2]],
            scale = c(4, .5),
            random.order = FALSE,
            rot.per = 0, # % of vertical words
            min.freq = 2,
            max.words = Inf,
            colors = brewer.pal(8, "Dark2")[-(1:2)])
  title(names(ldf[i]), line = 2) # position of the title
}
```

Remove temporal wordcloud data

```{bash}
rm -r wordclouds
```

### Wordclouds by terms

```{r, fig.show = "hold", out.width = "50%", fig.height=6, fig.width=6}
for (i in 1:length(enrichment_list[[LIST_LEVEL]])) {
  wcdf <- read.table(text = enrichment_list[[LIST_LEVEL]][[i]]$GeneRatio,
                     sep = "/")[1]
  names(wcdf)[1] <- "GeneRatio"
  # add GO term column
  wcdf$term <- enrichment_list[[LIST_LEVEL]][[i]][,2]
  # calculate MIN.FREQ
  MIN_FREQ <- max(wcdf$GeneRatio)/MIN_RATIO
  set.seed(SEED)
  wordcloud(words = wcdf$term,
            freq = wcdf$GeneRatio,
            scale = c(.8, .3),
            random.order = FALSE,
            rot.per = 0, # % of vertical words
            min.freq = MIN_FREQ,
            max.words = Inf,
            colors = brewer.pal(8, "Dark2")[-(1:2)])
  title(names(enrichment_list[[LIST_LEVEL]][i]), line = 2)
}
```

### Convert IDs Arabidopsis to original organism and save (if the orthology was performed)

```{r, eval=MODEL_REFERENCE==FALSE, include=MODEL_REFERENCE==FALSE}
# create dataframe to save data
enriched_genes <- list()
for (i in 1:length(enrichment_list[[LIST_LEVEL]])) {
  tmp <- enrichment_list[[LIST_LEVEL]][[i]]@result
  # filter results to keep only significant
  # because in @result all the genes are saved
  tmp <- tmp[tmp$p.adjust <= P_CUTOFF, ]
  tmp <- tmp[tmp$qvalue <= Q_CUTOFF,]
  tmp <- tmp[, c(2, 8)]
  name <- names(enrichment_list[[LIST_LEVEL]])[i]
  final <- data.frame()
  for (j in 1:nrow(tmp)) {
    tmp2 <- tmp[j, 2]
    tmp2 <- as.data.frame(t(read.table(text = tmp2, sep = "/")))
    tmp2$process <- tmp[j,1]
    colnames(tmp2)[1] <- ORG_ENRICH
    final <- rbind(final, tmp2)
  }
  # merge with genes_list_arab to get the orthology
  tmpy <- genes_list_arab[[name]]
  tmpy <- tmpy[, c(1, 2)]
  tmp3 <- merge(x = final, y = tmpy,
                by.x = 1, by.y = 2,
                all = FALSE)
  tmp3 <- tmp3[order(tmp3$process, decreasing = FALSE), ]
  # save in the list
  enriched_genes[[name]] <- tmp3
}
save_enriched_genes <- as.data.frame(do.call(rbind, enriched_genes))
save_enriched_genes$condition <- rownames(save_enriched_genes)
save_enriched_genes$condition <- gsub("\\.[0-9].*", "", save_enriched_genes$condition)
# reorder dataframe
save_enriched_genes <- save_enriched_genes[, c(4, 1, 3, 2)]
write.table(save_enriched_genes,
            file = paste0(SUB_DIR,
                          LIST_LEVEL,
                          "_genes_arabidopsis_picual.tsv"),
            quote = FALSE,
            sep = "\t",
            row.names = FALSE)

# remove useless variables
rm(tmp, name, final, tmp2, tmpy, tmp3, enriched_genes, save_enriched_genes)
```


